```{r cache = T, echo = F, message=F}
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(fig.path = "images/finding-stage-pace-")
```
# Finding Pace Across Stages

Average speed on a rally is all very well, but it's not the most useful of metrics for making sense of what's actually going on in a rally. Far more useful is the notion of *pace* the reciprocal of a speed like measure, that tells you how many seconds it's taking each driver to cover one kilometer.

Knowing the pace allows you to make more direct comparisons between drivers, as well as simplifying rule of thumb calculations, like what sort of pace advantage a driver needs to make up the 2s to the leader over the remaining 100 kilometers available in the final four stages...

In this chapter, we look at some simple pace calculations, rebase pace values relative to a specified driver, and explore a couple of ways of visualising differential pace over the course of a rally in the form of *pace maps* and *off-the-pace* charts.

## Load Base Data

To get the stage data from a standing start, we can load in the current season list, select the rally we want, look up the itinerary from the rally, extract the sections and then the stages, and from that access the stage ID for the stage or stages we are interested in.

Load in the helper functions:

```{r message=F, warning=F}
source('code/wrc-api.R')
```

And get the base data:

```{r}
s = get_active_season()
eventId = get_eventId_from_name(s, 'arctic')

itinerary = get_itinerary(eventId)
sections = get_sections(itinerary)
stages = get_stages(sections)
stages_lookup = get_stages_lookup(stages)

# Driver details
entries = get_rally_entries(eventId)
cars = get_car_data(entries)
```

Get a sample stage ID:

```{r}
stageId = stages_lookup[['SS3']]
```

## Defining Pace

With variable stage distances on a stage rally, metrics such as average speed provide one way of comparing performances across stage, calculated as $\textrm{stage_time}/\textrm{stage_distance}$ with units of *kilometers* or *miles per hour*.

A more useful measure, particularly in rally terms, is the notion of *pace*, typically given with units of *seconds per kilometer*. *Speed* tells us much quickly a car covers distance in unit time; *pace* gives us an indication of how much time is required to travel a unit distance.

When used as a rebased difference measure between drivers, pace difference allows us to rapidly calculate how much time a driver is likely to gain or lose over a particular stage distance as per the word equation $\textrm{time_gain}=\textrm{stage_distance}\cdot\textrm{pace_difference}$.

Basic pace itself is given as $\textrm{pace}=\textrm{time}/\textrm{distance}$.

Developing our rally algebra, we might identify the stage distance for stage $S$ as $d_S$. For a stage time by driver $i$ of ${_S}t_i$ the stage pace ${_S}p_i$ for driver $i$ on stage $S$ is then given as:

$$
{_S}p_i = \frac{{_S}t_i}{d_S}
$$

## Calculating Stage Pace

We can calculate stage pace from stage times and stage distances.

We can find stages distances directly from the `stages` dataframe:

```{r}
stages %>% select(c('code','distance')) %>% head(3)
```

### Calculating Pace for a Single Stage

Let's start by looking a single stage using a recipe we have used before:

```{r}
# Example stage code
stage_code = 'SS3'

stageId = stages_lookup[[stage_code]]

# Get the stage distance
stage_distance = stages[stages['code']==stage_code, 'distance']

# Get driver metadata
cars = get_car_data(entries)

# Create stage times with merged in driver metadata
stage_times = get_stage_times(eventId, stageId) %>%
                      arrange(position) %>%
                      head(10) %>%
                      # Merge in the entries data
                      merge(cars, by='entryId')  %>%
                      # Convert milliseconds to seconds
                      mutate(TimeInS = elapsedDurationMs/1000)  %>%
                      # Limit columns and set column order
                      select(c('position', 'identifier',
                               'code', 'TimeInS')) %>%
                      # The merge may upset the row order
                      # so reset the order again
                      arrange(position) %>%
                      # Improve column names by renaming them
                      rename(Pos=position,
                             Car = identifier,
                             Code = code,
                             `Time (s)` = TimeInS)

formattable(stage_times )
```

We can now calculate pace as the stage time divided by the stage distance:

```{r}
stage_times$pace = stage_times$'Time (s)' / stage_distance

stage_times
```

### Calculating Pace for Multiple Stages

First, let's get the data for all the stages:

```{r}
stage_list = get_stage_list(stages)

multi_stage_times = get_multi_stage_times(stage_list)
  
multi_stage_times %>% tail(2)
```

We can generate the pace by adding the stage distance as an extra column and performing the pace calculation.

We'll also take the opportunity to merge in driver metadata and limit cars to WRC group entries:

```{r}
get_multi_stage_pace = function(multi_stage_times, cars) {
  multi_stage_times %>%
                    merge(stages[,c('stageId' ,'distance',
                                    'number', 'code')],
                          by='stageId') %>%
                    mutate(elapsedDurationS = elapsedDurationMs / 1000,
                            pace = elapsedDurationS / distance) %>%
                    merge(cars[,c('entryId','drivername',
                                  'code', 'groupname')],
                                    by='entryId',
                          suffixes=c('','_driver')) %>%
                    filter(groupname=='WRC') %>%
                    select(c('stageId', 'number', 'code_driver',
                             'elapsedDurationS', 'pace', 'code'))  %>%
                    arrange(number, elapsedDurationS)
  
}

multi_stage_pace = get_multi_stage_pace(multi_stage_times, cars)

multi_stage_pace %>% head(3)
```

Create a mapping from stage ID to stage codes and cast the ordered list of stage Ids to an ordered list of stage codes:

```{r}
get_stage_codes = function(stages){
     # Create a stage code mapping function
    stages_lookup_code = get_stages_lookup(stages, 'stageId', 'code')
    stage_code_map = function(stageId)
      stages_lookup_code[[as.character(stageId)]]
    
    # Map stage ID column names to stage codes
    stage_codes = unlist(purrr::map(stage_list,
                                    function (x) stage_code_map(x)))
    stage_codes 
}

stage_codes = get_stage_codes(stages)
```

Use the generic widener function to widen the pace dataframe to give the pace for each driver on each stage:

```{r}
pace_wide = get_multi_stage_generic_wide(multi_stage_pace,
                                         stage_codes, 'pace',
                                         # Unique group keys required
                                         # Driver code not guaranteed unique
                                         group_key=c('code_driver'),
                                         spread_key='code')

pace_wide %>% head(3)
```

## Rebasing Stage Pace

We can rebase the stage pace according to a specific driver:

```{r}
example_driver = pace_wide[2,]$code_driver

pace_wide_rebased = rebase(pace_wide, example_driver, stage_codes,
                           id_col='code_driver')

pace_wide_rebased %>% head(3)
```

More abstractly, the rebased pace, ${_S}p_i^j$, for driver $i$ relative to driver $j$ on stage $S$ is given as:

$$
{_S}p_i^j = {_S}p_i - {_S}p_j = \frac{{_S}t_i - {_S}t_j}{d_S} = \frac{{_S}t_i^j}{d_S}
$$

