
Although the `formattable()` function does not directly support  colour bars that change color, and direction, depending on whether the associated value is positive or negative, we can create a custom formatter that does provide such a view over the data.
As well as left and right directed coloured bars,

For example, if we use our rebased data set, we can highlight the positive and negative gaps to each other driver:

```{r}
# https://github.com/renkun-ken/formattable/issues/95#issuecomment-364459116
pm_color_bar <- function(color1 = "lightgreen", color2 = "pink", ...){
  formatter("span",
            style = function(x) style(
              display = "inline-block",
              float = ifelse(x >= 0, "right", "left"),
              "text-align" = ifelse(x >= 0, "right", "left"),
              "margin-left" = ifelse(x >= 0, "0%", "50%"),
              "margin-right" = ifelse(x >= 0,"50%", "0%"),
              "border-radius" = "4px",
              "background-color" = ifelse(x >= 0, color1, color2),
              width = percent(0.5*proportion(abs(as.numeric(x)), ...))
            ))
}

formattable(rov_rebased_gap,
            list(Gap = pm_color_bar()))
```

A problem with this is that the width of the cell used to colour the bar also sets the text width. Instead, we need to separate the background from the text:

```{r}
bg = function(x, xnorm, color, ...) {
  if (color=='pink'){
    start = percent(0)
    mid = percent(xnorm)
    end = percent(0.5)
  } else {
    start=percent(0)
    mid = percent(0.5)
    end = percent(xnorm)
  }
  #paste("linear-gradient(90deg,", color, pc,", transparent ",pc,")")
  #paste0("linear-gradient(90deg,transparent ", start,"%,", color," ",start,"%, color ",end,"%, transparent")
  #paste0("linear-gradient(90deg,transparent ", 0,"%", color," ",50,"% ",color,  75,"% transparent")
  #paste("linear-gradient(90deg,", color, pc,", transparent ",pc,")")
  #paste0("linear-gradient(90deg,transparent ",start,",", color, " ",start,", transparent ",end,")")
  paste("linear-gradient(90deg,transparent ",start,",",
        color, mid,
        ", transparent", end,")")
} 
```
save to png

(https://stackoverflow.com/questions/38833219/command-for-exporting-saving-table-made-with-formattable-package-in-r)

https://rdrr.io/cran/rbokeh/man/widget2png.html

https://rdrr.io/cran/htmlwidgets/man/saveWidget.html

https://www.r-graph-gallery.com/159-save-interactive-streamgraph-to-static-image-png.html


https://github.com/renkun-ken/formattable/issues/26


```{r}
pm_color_bar2 <- function(color1 = "lightgreen", color2 = "pink", ...){
  formatter("span",
            style = function(x) style(
              display = "inline-block",
              #"text-align" = 'center',
              "border-radius" = "4px",
              # We need to use the linear gradient approach
              #"background" = "linear-gradient(90deg,#d65f5f 89.2%, transparent 89.2%)"#,
              "width"='10em',
              "background" = bg(x, xnormalize(x), ifelse(x >= 0, color1, color2))
            ))
}

formattable(rov_rebased_gap,
            list(Gap = pm_color_bar2()))
```

  # Get stage codes lookup id->code
  stages_lookup_code = get_stages_lookup(stages, 'stageId', 'code')
  
  # Map stage ID column names to stage codes
  stage_codes = unlist(purrr::map(stage_list,
                                  function (x)
                                    stages_lookup_code[[as.character(x)]]))

  colnames(df) = c('entryId', stage_codes)