```{r cache = T, echo = F, message=F}
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(fig.path = "images/itinerary-")
```
# Itinerary & Road Position

The competitive phase of a full WRC Rally event typically extends over three days (Friday to Sunday), with either a ceremonial start or a short first stage on the Thursday evening. Shorter format events are also possible.

Each day is referred to as a *leg*, and each leg is structured as a *section*, often referred to as a *loop*.

## Load Base Data

Start by loading in the base WRC API helper functions:

```{r message=F, warning=F}
source('code/wrc-api.R')
library(tidyr)
```

And grab some minimal event metadata:

```{r}
s = get_active_season()
eventId = get_eventId_from_name(s, 'arctic')
```

## Displaying the Itinerary

We can grab the full itinerary with a single function call:

```{r}
itinerary = get_itinerary(eventId)
itinerary
```

The *status* often does not get updated at the end of the event, so completed events may still describe the final day as *Running*.

```{r}
itinerary %>% select(c('name', 'legDate'))
```

### Itinerary Legs

Let's have a look at the structure of a particular day:

```{r message=FALSE, warning=FALSE}
example_section = itinerary[[1, 'itinerarySections']]
example_section
```

We can also get the full set of itinerary sections in one dataframe:

```{r}
itinerary_sections_full = do.call(rbind, itinerary$itinerarySections)

itinerary_sections_full
```

### Itinerary Controls

The *controls* column details information about all the timing controls:

```{r}
example_controls = example_section[[1, 'controls']]
example_controls
```

Let's see the key information for the controls:

```{r message=FALSE}
controls_cols = c('controlId', 'eventId', 'type',
                  'code', 'location', 'distance', 'firstCarDueDateTime')

example_controls %>% select(controls_cols)
```

We can get a list of all the controls by combining data from the separate legs dataframes:

```{r}
get_multi_controls = function(itinerary_sections){
  multi_controls = do.call(rbind, itinerary_sections$controls)

  multi_controls %>% select(controls_cols)
}
```

Let's see how it works:

```{r}
multi_controls = get_multi_controls(itinerary_sections_full)
multi_controls %>% tail()
```

### Itinerary Stages

The *stages* column provides more information about each stage:

```{r}
example_stages = example_section[[1, 'stages']]
example_stages
```

Let's focus on the key columns:

```{r meddage=FALSE}
stages_cols = c('stageId', 'eventId', 'number', 'name',
                'distance', 'stageType', 'code')

stage_name_cleaner = function(df) {
  df %>%
    mutate(fullname=name,
         name=stringr::str_replace(name, ' \\(Live TV\\)', '')) %>%
    mutate(fullname=name,
         name=stringr::str_replace(name, ' \\(Wolf Power Stage\\)', ''))
}

example_stages %>% 
  select(stages_cols) %>%
  stage_name_cleaner
```

Once again, we can pull all the information into a single dataframe:

```{r}
get_multi_stage_details = function(itinerary){
  multi_stage_details = do.call(rbind, itinerary$stages)

  multi_stage_details %>% 
    select(stages_cols) %>%
    stage_name_cleaner()
}
```

Here's how it works:

```{r}
multi_stage_details = get_multi_stage_details(itinerary_sections_full)
multi_stage_details %>% tail()
```



