```{r cache = T, echo = F, message=F}
knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(fig.path = "images/multi-stage-results-")
```
# Visualising Results for Multiple Stages

As well as visualising the results for a single stage, we might want to visualise the results over multiple stages. The basic overall results can be retrieved from a single call to the WRC results API, but to view the stage times and rankings across multiple stages requires retrieving detailed for each stage and then combining it into a single dataframe.

## Load Base Data

To get the splits data from a standing start, we can load in the current season list, select the rally we want, look up the itinerary from the rally, extract the sections and then the stages and the retrieve the stage ID for the stage we are interested in.

To begin with, load in our WRC API helper functions:

```{r message=F, warning=F}
source('code/wrc-api.R')
```

Now let's grab some data:

```{r}
s = get_active_season()
eventId = get_eventId_from_name(s, 'arctic')

entries = get_rally_entries(eventId)

itinerary = get_itinerary(eventId)
sections = get_sections(itinerary)
stages = get_stages(sections)
stages_lookup = get_stages_lookup(stages)

stage_list = get_stage_list(stages)
stage_codes = stages$code
# To generate stage codes as an ordered factor:
# factor(stages$code, levels = stages$code)
```


## Retrieving Stage Results for Multiple Stages

To being with, lets get the overall results at the end of each stage:

```{r}
multi_overall_results = get_multi_overall(stage_list)

multi_overall_results  %>% tail(2)
```

### Widening Overall Position

We can reduce the amount of data by casting the long raw result to a wide format, widening the data on a particular field of interest. For example, we can widen generate a wide dataframe describing overall positions, ${_S}o$ at the end of each stage, where a particular driver's position is given as ${_S}o_i$:

```{r}
multi_overall_wide_pos = multi_overall_results %>%
                            get_multi_stage_generic_wide(stage_list,
                                                         'position')

multi_overall_wide_pos %>% head(2)
```

### Widening Overall Rally Time

We can also create a wide format report of the overall times, where each column gives the overall, accumulated rally time up to and including each stage, ${_S}T$; each cell then represents the accumulated time for a particular driver, ${_S}T_i$.

The times themselves appear in units of milliseconds, so first create a column corresponding to time in seconds, then widen using those values:

```{r}
multi_overall_results = multi_overall_results %>%
                            mutate(totalTimeS = totalTimeMs/1000)

multi_overall_wide_time = multi_overall_results %>%
                              get_multi_stage_generic_wide(stage_list,
                                                           'totalTimeS')

multi_overall_wide_time %>% head(2)
```

We note that with stages presented in order, the rally time is strictly increasing across the rows.

We further note that we can derive stage times from the overall rally times by calculating the columnwise differences ${_S}t={_S}T-{_{S-1}}T: 1<S<N$ for an $N$ stage rally.

### Widening Time to First

Another useful time is the *time to first*, which is to say, the *gap*, ${_S}GAP_i$. Noting that the overall rally leader may change at the end of each stage, this measure is essentially a rebasing measure relative to a particular position rather than a particular driver:

```{r}
multi_overall_results = multi_overall_results %>%
                            mutate(diffFirstS = diffFirstMs/1000)

multi_overall_wide_gap = multi_overall_results %>%
                            get_multi_stage_generic_wide(stage_list,
                                                         'diffFirstS')

multi_overall_wide_gap %>% head(2)
```

However, we could also calculate the gap to leader from the overall times, first by identifying the minimum accumulated time at each stage (that is, the minimum time, excluding null values, in each overall time column) and then by subtracting those values from each row in the overall times dataframe, which is to say:

$$
\textrm{GAP}_{i} = {_S}t_i - \textrm{min}({_S}T)
$$

If we try to subtract a list of values (for example, $\forall s \in [1 \le S \le N]: \textrm{min}({_S}T)$) from an *R* dataframe, we need to tell *R* how we want that subtraction performed. Internally, the dataframe is represented as a long list of values made up from values in the first column, then the second, and so on. If we subtract a list of *N* values from the dataframe, the values are selected from the first *N* items in this long serialised version of the dataframe, then the next *N* values and so on.

So to subtract a "dummy" row of values from the dataframe, we need another approach. The `purrr::map2df()` function allows us to apply a function, in this case the subtraction `-` function, with a set of specified values we want to subtract, from each row in the dataframe.

So let's create a set of values representing the minimum overall time in each stage. The `matrixStats::colMins()` function will find the minimum values by row from a matrix, so cast the stage time columns from the wide dataframe to an appropriately sized matrix and then find the minimum in each column, ignoring null values:

```{r}
overall_m = as.matrix(multi_overall_wide_time[,as.character(stage_list)],
                      ncols=length(stage_list))

mins_overall = matrixStats::colMins(overall_m, na.rm=TRUE)

mins_overall
```

We can now subtract this "dummy row" of values from each row in the dataframe to find the gap to leader for each row on each stage:

```{r}
purrr::map2_df(multi_overall_wide_time[,as.character(stage_list)], 
              mins_overall, `-`) %>% head(2)
```

Comparison with the "diffToFirst" times should show them to be the same.

### Mapping Stage and Driver Identifiers to Meaningful Labels

To improve the look of the table, we might use stage codes and driver codes to label the columns and identify the rows.

To start with, we can map the column names that correspond to stage codes via a lookup list of stage ID to stage code values:

```{r}
map_stage_codes = function(df, stage_list) {
  # Get stage codes lookup id->code
  stages_lookup_code = get_stages_lookup(stages, 'stageId', 'code')
  
  #https://stackoverflow.com/a/34299333/454773
  plyr::rename(df, replace = stages_lookup_code,
               warn_missing = FALSE)
}

multi_overall_wide_time = multi_overall_wide_time %>%
                            map_stage_codes(stage_list)

multi_overall_wide_time %>% head(2)
```

We can also create a function to replace the entry ID with the driver code.
Note also the select statement at the end that puts the columns into a sensible order:

```{r}
cars = get_car_data(entries)

map_driver_names = function(df, cars){
  df %>%
    merge(cars[,c('entryId','code')],
          by='entryId')  %>%
    # Limit columns and set column order
    select(-'entryId') %>%
    # Move last column to first
    select('code', everything())
}

multi_overall_wide_time = multi_overall_wide_time %>%
                            map_driver_names(cars)

multi_overall_wide_time %>% head(2)
```

### Rebasing Overall Times

We can rebase the overall times with respect to a particular driver in the normal way:

```{r}
example_driver = multi_overall_wide_time[2,]$code

overall_wide_time_rebased = rebase(multi_overall_wide_time,
                                   example_driver, stage_codes,
                                   id_col='code')

overall_wide_time_rebased %>% head(3)
```

### Finding Changes in Rebased Gaps Across Stages

The rebasing operation essentially allows us to select a row of times for one particular driver and then subtract that row from every other row to give us a direct comparison of the gap between a specified driver and every other driver.

But we can also perform a consecutive column-wise differencing operation on the rebased times that allows to see how much time was gained or relative to a particular driver in going from one stage to the next (observant readers may note that this results in the rebased stage time for each stage...).

To subtract one column from the next, create two offset dataframes, one containing all but the first stage (first stage column) and one containing all but the last stage (final stage column). If we subtract one dataframe from the other, it gives us our column differences. Inserting the original first column back in its rightful place gives us the columnwise differences table:

```{r}
#https://stackoverflow.com/a/50411529/454773
df = overall_wide_time_rebased

# [-1] drops the first column, [-ncol()] drops the last
 df_ = df[,stage_codes][-1] - df[,stage_codes][-ncol(df[,stage_codes])]
  
# The split time to the first split is simply the first split time
df_[stage_codes[1]] = df[stage_codes[1]]
# Return the dataframe in a sensible column order
df_ %>% select(stage_codes) %>% head(3)
```

*(A similar technique could be used to recreate stage times from the overall times.)*

## Visualising Multiple Stage Results

The overall stage results provides information regarding the overall times and positions at the end of each stage; the rebased overall times provide us with gap information from a specified driver to every other driver.

So how might we use exploratory data visualisation techniques to support a conversation with that data or highlight potential stories hidden within it?

### Visualising First Position

One way of enriching the wide position table might be to highlight the driver in first position at the end of each stage. We can do this using the `formattable::formattable()` function.

First, let's tidy up the overall position table:

```{r}
multi_overall_wide_pos = multi_overall_wide_pos %>%
                            map_stage_codes(stage_codes) %>%
                            map_driver_names(cars)

multi_overall_wide_pos %>% head(2)
```
We can also reorder the table by the final stage position, as described by the last stage code in the `stage_codes` list. However, because we want to sort on a column name as provided by a variable, we need to use the `!!` operator to force the evaluation of the single variable value to the column name symbol (`as.symbol()`):

```{r}
multi_overall_wide_pos = multi_overall_wide_pos %>%
              dplyr::arrange(!!as.symbol(stage_codes[length(stage_codes)]))

multi_overall_wide_pos %>% head(2)
```

Create a function to highlight the first position car:

```{r}
library(formattable)

highlight_first =  function (...) 
{
  formatter("span",
            style = function(x) ifelse(x==1,
                                       style(display = "block", 
                                             padding = "0 4px",
                                             `color` = "black",
                                             `column-width`='4em',
                                             `border-radius` = "4px",
                                             `background-color` = 'lightgrey'),
                                      style()))
}
```

And then use that function to help style the table:

```{r}
multi_overall_wide_pos %>%
      head(3) %>%
      formattable(# Align values in the center of each column
                  align='c',
                  list(area(col = stage_codes) ~ highlight_first()))
```

### Visualising Position Changes Over a Rally

Another useful way of summarising the positions is to chart showing the evolution of position changes.

The chart is constructed most straightforwardly from tidy long format) data:

```{r}
overall_pos_long_top10 <- multi_overall_wide_pos %>%
                              head(10) %>%
                              gather(key ="Stage",
                                     value ="Pos",
                                     stage_codes)

overall_pos_long_top10 %>% head(3)
```

It will be convenient for the stage codes to be represented as ordered factors:

```{r}
overall_pos_long_top10 = overall_pos_long_top10 %>%
                             mutate(Stage = factor(Stage,
                                                   levels = stage_codes))

overall_pos_long_top10$Stage[1]
```

```{r simple_pos_chart, message=FALSE, warning=FALSE}
library(ggplot2)

pos_range = 1:max(overall_pos_long_top10$Pos)
g_pos = ggplot(overall_pos_long_top10, aes(x=Stage, y=Pos)) +
                  geom_line(aes(group = code)) +
                  # Invert scale and relabel y-axis
                  # https://stackoverflow.com/a/28392170/454773
                  scale_y_continuous(trans = "reverse",
                                     breaks = pos_range) +
      theme_classic()

g_pos
```

We can producing a cleaner chart by adding driver labels to the start and end of each line using the `directlabels:geom_dl()` function, as well as dropping the axes:

```{r cleaner_pos_chart, message=FALSE, warning=FALSE}
library(directlabels)

g_pos +
    geom_dl(aes(label = paste0(' ',code)), # Add space before label
            # cex is text label size
            method = list('last.bumpup', cex = 0.5)) +
    geom_dl(aes(label = paste0(code, ' ')), # Add space before label
            method = list('first.points', cex = 0.5)) +
    theme_void()
```

In the above chart, you may notice a gap at first stage position 6 where Ogier was originally placed. A more robust way to prepare the data for this sort of charts is to filter the data by class, for example limiting the data to cars in the *WRC* group/class, and then reranking the position by group/class. A finished position chart can then position the drivers by class ranking and use labels to overplot actual overall rally positions on stages where the overall stage position differs from the class rank.

### Visualising Position Gains/Losses

To visualise position changes for a driver from one driver to the next, we can create a table of position differences. Let's abstract out the code we used to find differences between columns in to a function:

```{r}
coldiffs = function(df, cols, dropfirst=FALSE, firstcol=NULL){
  cols = as.character(cols)
  # [-1] drops the first column, [-ncol()] drops the last
  df_ = df[,cols][-1] - df[,cols][-ncol(df[,cols])]
  
  # The split time to the first split is simply the first split time
  df_[cols[1]] = df[cols[1]]
  # Return the dataframe in a sensible column order
  df_ = df_ %>% select(cols)
  
  if (!is.null(firstcol))
    df_[, cols[1]] = firstcol
  
  if (dropfirst)
    df_[,cols][-1]
  else
    df_
}
```

Let's put that function through its paces. First, we can drop the first column:

```{r}
coldiffs(multi_overall_wide_pos, stage_codes, dropfirst=TRUE) %>% head(2)
```

Then we can retain the first column and replace it with a specified value:

```{r}
coldiffs(multi_overall_wide_pos, stage_codes, firstcol=999) %>% head(2)
```

We can now fund the position changes from stage to stage as well as tidying up the identifiers:

```{r}
pos_diffs = multi_overall_wide_pos %>% coldiffs(stage_codes,
                                                firstcol=0)
pos_diffs$code = multi_overall_wide_pos$code

# Reorder the columns by moving last column to first
pos_diffs = pos_diffs %>% select('code', everything())

pos_diffs %>% head(3)
```

Note that we need to be careful with the sense of how we read this table: a *negative* position change means the driver has *improved* their position. It might be more meaningful to have position gain/loss, rather than strict position difference columns, where a positive value denotes an improved position:

```{r}
pos_gains = pos_diffs
pos_gains[,stage_codes] = -pos_gains[,stage_codes[-1]]

pos_gains %>% head(3)
```

One  way of highlighting position changes is to use coloured up/down arrows:

```{r}
updown = function(...){
  formatter("span", 
            style = function(x) style(color = ifelse(x>0,
                                                     "green",
                                                     ifelse(x<0,
                                                            "red",
                                                            "lightgrey"))),          
            function(x) icontext(ifelse (x >0,
                                         # i.e. gained position
                                         "arrow-up",
                                         ifelse (x < 0,
                                                 "arrow-down" ,
                                                 "resize-horizontal"))))
}
```

Let's see how that works (note that we need to cast the table `as.htmlwidget()` in order to render the arrows appropriately :

```{r}
pos_gains %>%
  head(3) %>% 
  formattable(list( 
    area(col = stage_codes) ~ updown())) %>%
  as.htmlwidget()
```

We can extend the formatter to also display the number of positions gained or lost:

```{r}
updown2 = function(...){
  formatter("span", 
            style = function(x) style(color = ifelse(x>0,
                                                     "green",
                                                     ifelse(x<0,
                                                            "red",
                                                            "lightgrey"))),          
            function(x) icontext(ifelse (x >0,
                                         # i.e. gained position
                                         "arrow-up",
                                         ifelse (x < 0,
                                                 "arrow-down" ,
                                                 "resize-horizontal")),
                                 # Add in the pos change value
                                 ifelse (x!=0, paste0('(',abs(x),')'),'')))
}
```

Let's see how it looks:

```{r}
pos_gains %>%
  head(3) %>% 
  formattable(list( 
    area(col = stage_codes) ~ updown2())) %>%
  as.htmlwidget()
```

Another way of visualising position changes is to create a simple summarising sparkline using the `sparkline::spk_chr()` function.

This requires us first to cast the data into a long format:

```{r}
pos_gain_long_top10 <- pos_gains %>%
                              head(10) %>%
                              gather(key ="Stage",
                                     value ="PosChange",
                                      stage_codes)

pos_gain_long_top10 %>% head(3)
```

We can then generate sparklines showing position changes:

```{r}
library(sparkline)

pos_gain_sparkline_top10 <- pos_gain_long_top10 %>%
                                group_by(code) %>%
                                summarize(spk_ = spk_chr(PosChange,
                                                         type ="bar"))

# We need to create an htmlwidget form of the table
out = as.htmlwidget(formattable(pos_gain_sparkline_top10))

# The table also has a requirement on the sparkline package
out$dependencies = c(out$dependencies,
                     htmlwidgets:::widget_dependencies("sparkline",
                                                       "sparkline"))
out
```

*Note that to render the sparkline, we need to cast the formatted table to an *htmlwidget* and also ensure that the required *sparkline* Javascript package is loaded into the widget.*

One issue with the sparkline bars is that the scales may differ. For example, across different drivers, a position change of +1 for one driver may have the same height as a position change of +2 for another driver.

## Retrieving Stage Times for Multiple Stages

This far, we have focused on exploring the overall results data. If required, we can also retrieve detailed results for multiple stages by requesting stage results for a specified list of stages:

```{r}
multi_stage_times = get_multi_stage_times(stage_list)
  
multi_stage_times %>% tail(2)
```

We can then cast the data to a wide format and relabel the resulting dataframe. For example, we can get a wide format dataframe containing individual stage times back:

```{r}
multi_stage_times_wide = multi_stage_times %>%
                            get_multi_stage_times_wide(stage_list) %>%
                            map_stage_codes(stage_list)  %>%
                            map_driver_names(cars)

multi_stage_times_wide %>% head(2)
```

Or we can get the stage rankings for each stage:

```{r}
multi_stage_wide_pos = multi_stage_times %>%
                            get_multi_stage_generic_wide(stage_list,
                                                         'position') %>%
                            map_stage_codes(stage_list)  %>%
                            map_driver_names(cars)

multi_stage_wide_pos %>% head(2)
```





